---
title: "Analysis of NKI mouse data"
subtitle: 'Analysis demo'
author: "Developed by [Gabriel Hoffman](http://gabrielhoffman.github.io/)"
date: "Run on `r Sys.time()`"
documentclass: article
output: 
  html_document:
  toc: true
  smart: false
---


<!--- 
rmarkdown::render("mouse_NKI.Rmd");

--->



```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  dev = c("png", "pdf"),
  package.startup.message = FALSE,
  cache = TRUE,
  cache.lazy = FALSE)
```

```{r load.packages}
library(ggplot2)
library(variancePartition)
library(edgeR)
library(kableExtra)
library(EnrichmentBrowser)
library(GSEABase)
library(zenith)
library(tidyverse)
```

# Read data and perform basic analysis
```{r load.data}
countMatrix = read.table("~/Downloads/counts.txt", header=TRUE, row.names=1, check.names=FALSE)
info = read.csv("~/Downloads/samples_ch+.csv", header=TRUE, row.names=1)

# filter genes by number of counts
isexpr = rowSums(cpm(countMatrix)>0.1) >= 10

# Standard usage of limma/voom
geneExpr = DGEList( countMatrix[isexpr,] )
geneExpr = calcNormFactors( geneExpr )

# estimate precision weights 
# There are 4 Groups.  
# By default the formula ~ Group will fit an intercept and 3 coefficients showing deviation from this intercept.
# But in your application, its simpler to fit all 4 Groups directly removing the intercept term with a 0 +:
form = ~ 0 + Group
vobj = voomWithDreamWeights( geneExpr, form, info, plot=TRUE )
```

### Multidomensional scaling, like PCA
```{r mds}
dcmp = cmdscale(dist(t(vobj$E)))

df = data.frame(dcmp, info)

ggplot(df, aes(X1, X2, color=Group)) +
  geom_point() +
  theme_classic() +
  theme(aspect.ratio=1)
```


# dream with contrasts
You can include covariates here, but that will interfere with the interpretation of the estimated values.  That is why your baseline expression values didt look right before.  Covariate are good for the regression model, but you should report the direct baseline expression values
```{r dream}
# fit regression model with contrasts
### WARNING: Please rewrite these contrasts for your biological question.
### I just focused on the implementation here
L = makeContrastsDream( form, info, 
  contrasts = c( Treatment = "GroupTs_ChP - GroupTs_ChN",
  				 Controls = "Group2N_ChP - Group2N_ChN",
  	all4 = "GroupTs_ChP - GroupTs_ChN - (Group2N_ChP - Group2N_ChN)"))

plotContrasts(L) 

fit = dream( vobj, form, info, L)
fit = eBayes(fit)

# Here are the computed group means
head(coef(fit))

# get p-values from 'all4' contrast
topTable(fit, coef="all4")
```


## Gene set analysis with zenith
```{r zenith}
# Mouse Gene Ontology
# (takes a few minutes the first time)
gs.go = getGenesets(org = "mmu", 
                      db = "go", 
                      gene.id.type = "ENSEMBL", 
                      return.type = "GeneSetCollection")

# modify gene set name to show description
gs.go_rename = lapply(gs.go, function(gs) {
        setName(gs) = paste0(setName(gs), ": ", description(gs))
        gs
    })
gs.go_rename = GeneSetCollection(gs.go_rename)

# run gene set analysis for each contrast
res_gsa = zenith_gsa( fit, gs.go_rename, coefs=colnames(L))
```

```{r top.genesets}
# examine top genesets for Treatment contrasts
res_gsa %>%
  filter(coef == "Treatment") %>%
  head %>%
  kbl(row.names=FALSE) %>% 
  kable_styling
```






#### SessionInfo
<details>
```{R sessionInfo}
sessionInfo()
```
<\details>
















