---
title: "Compare fixed vs mixed effects models"
subtitle: 'Simple null simulation'
author: "Developed by [Gabriel Hoffman](http://gabrielhoffman.github.io/)"
date: "Run on `r Sys.time()`"
documentclass: article
output: 
  html_document:
  toc: true
  smart: false
---




```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  dev = c("png", "pdf"),
  package.startup.message = FALSE,
  cache = TRUE,
  cache.lazy = TRUE)
```


```{r sim.dat}
library(PRROC)
library(lmerTest)
library(parallel)
library(mvtnorm)
library(ggplot2)
library(decorrelate)
RNGkind("L'Ecuyer-CMRG")

n = 1000
n_sims = 5000

beta_true = .05

# equal size batches
# n_per_batch = seq(2, n/2)
# value = n / n_per_batch
# n_batches = n / n_per_batch[value == floor(value)]

# uneven batches
n_batches = floor(seq(2, (n/2), length.out=100))

res = mclapply(n_batches, function(n_batches){

	# df = data.frame(Batch = factor(rep(rep(1:n_batches), n / n_batches)))
	df = data.frame(Batch = factor(sample.int(n_batches, n, replace=TRUE)))
	Z_mat = model.matrix( ~ 0 + Batch, df)

	set.seed(1)
	res = mclapply(1:n_sims, function(i){
		RhpcBLASctl::omp_set_num_threads(1)	

		beta = ifelse(i < n_sims/10, beta_true, 0)
			
		df$z = rnorm(n)
		df$x = df$z + rnorm(n)
		df$y = beta*df$x + 2*df$z + rnorm(n)

		df_local = df

		mod1 = coef(summary(lm(y ~ x, df_local)))
		mod2 = coef(summary(lm(y ~ x + z + Batch, df_local)))
		# mod3 = coef(summary(lmer(y ~ x + z + (1|Batch), df_local)))

		data.frame(i, beta, n_batches,
					mean_batch_size = mean(colSums(Z_mat)),
					'no covariates' = mod1[2,4], 
					`covariates (fixed)` = mod2[2,4], 
					# `covariates (mixed)` = mod3[2,5], 
					check.names=FALSE)
	}, mc.cores=12)
	do.call(rbind, res)
}, mc.cores=4)
res = do.call(rbind, res)



ids = colnames(res)[grep("covariates", colnames(res))]


df_perf = lapply( unique(res$n_batches), function(n_batches){
	resSub = res[res$n_batches == n_batches,]
	df_perf = lapply(ids, function(col){

		i = (resSub$beta != 0)
		pr = pr.curve(-log10(resSub[i,col]), -log10(resSub[!i,col]), rand.compute=TRUE, curve=TRUE)

		data.frame(n_batches, 
			mean_batch_size = resSub$mean_batch_size[1],
			Method=col, AUPR=pr$auc.integral, AUPR.rand = pr$rand$auc.integral)
	})
	do.call(rbind, df_perf)
})
df_perf = do.call(rbind, df_perf)

df_perf$Method = factor(df_perf$Method, rev(ids))
```

```{r plot1, cache=FALSE}
ggplot(df_perf, aes(n_batches, AUPR, color=Method)) +
	geom_point(alpha = .4) + 
	geom_smooth(se=FALSE) + 
	theme(aspect.ratio=1) +
	theme_classic() +
	scale_y_continuous(limits=c(0,1), expand=c(0,0)) +
	geom_hline(yintercept=df_perf$AUPR.rand[1], color="grey", linetype="dashed") +
	xlab("Number of batches")
```




```{r plot2, cache=FALSE}
ggplot(df_perf, aes(mean_batch_size, AUPR, color=Method)) +
	geom_point(alpha = .4) + 
	geom_smooth(se=FALSE) + 
	theme(aspect.ratio=1) +
	theme_classic() +
	scale_y_continuous(limits=c(0,1), expand=c(0,0)) +
	geom_hline(yintercept=df_perf$AUPR.rand[1], color="grey", linetype="dashed") +
	scale_x_log10() +
	xlab("Mean samples per batch")
```



```{r exit, cache=FALSE}
knitr::knit_exit()
```

cd /hpc/users/hoffmg01/work/misc_vp
ml git
git pull

system("git pull"); rmarkdown::render("test_mixed_model.Rmd")



res = mclapply(1:n_sims, function(i){

	beta = ifelse(i < n_sims/5, .16, 0)

	# Sig = autocorr.mat(3, .4)
	# Sig[2,1] = Sig[1,2] = beta
	# df = rmvnorm(n, c(0,0, 0), Sig)
	# colnames(df) = c('y', 'x', 'z')
	# df = data.frame(df)
	# cor(df)

	df$z = rnorm(n)
	df$x = .1*df$z + rnorm(n)
	df$y = beta*df$x + 5*df$z + rnorm(n)

	mod1 = coef(summary(lm(y ~ x, df)))
	mod2 = coef(summary(lm(y ~ x + z, df)))

	data.frame(i, beta,
					beta_est1 = mod1[2,1], p.1 = mod1[2,4], 
					beta_est2 = mod2[2,1], p.2 = mod2[2,4]) 
}, mc.cores=12)
res = do.call(rbind, res)



df_perf = lapply( c('p.1', 'p.2'), function(col){

	i = (res$beta != 0)
	pr = pr.curve(-log10(res[i,col]), -log10(res[!i,col]), rand.compute=TRUE, curve=TRUE)

	data.frame( col=col, AUPR=pr$auc.integral)
})	
df_perf = do.call(rbind, df_perf)


df_perf


ggplot(df_perf, aes(AUPR, color=col)) +
	geom_point() + 
	geom_line() + 
	theme(aspect.ratio=1) +
	theme_classic() +
	scale_y_continuous(limits=c(0,1), expand=c(0,0))






	Sig = matrix(.8, 2,2)
	diag(Sig) = 1 
	gamma = rmvnorm(ncol(Z_mat), c(0,0), Sig)







		library(RhpcBLASctl)
		RhpcBLASctl::omp_set_num_threads(1)



		# df_local$x = Z_mat %*% gamma[,1] 

		# eta_x = df_local$x

		# eta_z = Z_mat%*%gamma[,2]
		# # s = sqrt(var(eta_z) * (1-h_sq) / h_sq)
		# # eta_z = eta_z + rnorm(n, 0, s)
		# # cor(eta_x , eta_z)^2
		
		# eta = beta*eta_x + eta_z

		# s = sqrt(var(eta) * (1-h_sq) / h_sq)

		# df_local$y = eta + rnorm(n, 0, s)

		# cor(eta, df_local$y)^2



		mod1 = coef(summary(lm(y ~ x, df_local)))
		mod2 = coef(summary(lm(y ~ x + Batch, df_local)))
		# mod3 = coef(summary(lmer(y ~ x + (1|Batch), df_local)))

		# fit = lmer(y ~ x + (1|Batch), df_local)
		# calcVarPart(fit)
